---
output:
  pdf_document: default
  html_document:
    fig.caption: yes
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

# Small group assigment: Working with data in R

------------------------------------------------------------------------

Your task on this assignment is to work in a small group (2-3 people) to address the following questions. Only one writeup (as a .Rmd and .pdf) needs to be submitted for each group. The problems below use several different datasets. Your document should include code, figures, and written answers to each question.

## 1. Fisheries example

The code below includes the Canadian Atlantic cod population (in tonnes) over time. Build a plot of this data and explain in writing what has happened to the population.

```{r cod_pop,echo=T,eval=T, tidy=TRUE}
cod_pop <- c(1450, 1420, 1050, 2643, 1060, 1080, 1410, 1150, 740, 175, 43, 10, 12, 15, 16, 16, 28, 30, 32, 23, 12, 19, 27)
years <- c(1983, 1984, 1985, 1986, 1987, 1988, 1989, 1990, 1991, 1992, 
1993, 1994, 1995, 1996, 1997, 1998, 1999, 2000, 2001, 2002, 2003, 
2004, 2005)
```

```{r}
library(tidyverse)
cod_pop_over_time <- data.frame("Years" = years, 
                                "Cod Population"= cod_pop)

ggplot(data = cod_pop_over_time, aes(x= Years, y= Cod.Population))+
  geom_point()+
  ggtitle(label = "Cod Population Over Time")
```

This population of cod is going extinct over time, reaching nearly zero by 1994. 

Exponential growth can be used for populations that are still small or over short-time scales. The following code shows how a for loop can be used to build a plot of exponential growth.

$$N[t+1] = N[t] + rN[t]$$ where $N$ is the population size, $t$ is time, and $r$ is the instrinsic rate of growth
```{r}
##given in the newer qmd instructions on Canvas. 

{r}
#Step 1: Define parameters
r <- 0.46
t <- c(1:50)

#Step 2: make a vector to store population change
N <- vector(length = length(t))
N[1] <- 0.1 

#Step 3: run the for loop
for (i in t) {
  N[i+1] <-  N[i] + r*N[i]
}

plot(N, main = "Exponential Growth")
```

## 2-3. Population models

The discrete-time logistic growth model has the form $N[t] + r*N[t]*(1 - N[t]/K)$. It is often used to describe population growth, such as in animal and plant populations, cancer tumor growth, and bacterial growth. Your task is to write a for loop that runs this model from time = 1 to time = 50. The starting value of N is 0.1. The default parameter values are $r=0.4$ and $K=1$. What is the behavior (e.g., does the population go extinct? are there cycles?) with the default parameter values? What happens if you increase $r$? Please provide graphs to support your conclusions.

```{r}
#Initial loop with the default parameters. 
#Step 1: Define parameters
r <- 0.4
t <- c(1:50)
K<- 1

#Step 2: make a vector to store population change
N <- vector(length = length(t))
N[1] <- 0.1 


#Step 3: run the for loop
  for(i in t) {
  N[i+1] <-  N[i] + r*N[i]*((1-N[i])/K)
  }

plot(N, main= "r=0.4")


#Initial loop with the default parameters. 
#Step 1: Define parameters
r <- 0.5
t <- c(1:50)
K<- 1

#Step 2: make a vector to store population change
N <- vector(length = length(t))
N[1] <- 0.1 


#Step 3: run the for loop
  for(i in t) {
  N[i+1] <-  N[i] + r*N[i]*((1-N[i])/K)
  }

plot(N, main= "r=0.5")

#Initial loop with the default parameters. 
#Step 1: Define parameters
r <- 0.6
t <- c(1:50)
K<- 1

#Step 2: make a vector to store population change
N <- vector(length = length(t))
N[1] <- 0.1 


#Step 3: run the for loop
  for(i in t) {
  N[i+1] <-  N[i] + r*N[i]*((1-N[i])/K)
  }

plot(N, main= "r=0.6")

#Initial loop with the default parameters. 
#Step 1: Define parameters
r <- 0.7
t <- c(1:50)
K<- 1

#Step 2: make a vector to store population change
N <- vector(length = length(t))
N[1] <- 0.1 


#Step 3: run the for loop
  for(i in t) {
  N[i+1] <-  N[i] + r*N[i]*((1-N[i])/K)
  }

plot(N, main= "r=0.7")

#Initial loop with the default parameters. 
#Step 1: Define parameters
r <- 0.8
t <- c(1:50)
K<- 1

#Step 2: make a vector to store population change
N <- vector(length = length(t))
N[1] <- 0.1 


#Step 3: run the for loop
  for(i in t) {
  N[i+1] <-  N[i] + r*N[i]*((1-N[i])/K)
  }

plot(N, main= "r=0.8")

#Initial loop with the default parameters. 
#Step 1: Define parameters
r <- 0.9
t <- c(1:50)
K<- 1

#Step 2: make a vector to store population change
N <- vector(length = length(t))
N[1] <- 0.1 


#Step 3: run the for loop
  for(i in t) {
  N[i+1] <-  N[i] + r*N[i]*((1-N[i])/K)
  }

plot(N, main= "r=0.9")
```
With the default parameters N reaches K at ~20 generations. As the growth rate (r) increases N is able to reach K at a faster rate, and takes less generations to do so. This makes sense as with a faster growth rate you would expect a population to grow faster, and therefore reach it's capacity in less time. 

Given the graph above it appears that the population does not go extinct nor are there cycles. After about 20 

```{r}
#Lowering the growth rate to r = 0.1 so it takes longer than 20
#Step 1: Define parameters
r <- 0.05
t <- c(1:50)
K<- 1

#Step 2: make a vector to store population change
N <- vector(length = length(t))
N[1] <- 0.1 
#Step 2 pt2: make a vector to store the growth rate change


#Step 3: run the for loop
for (i in t) {
  N[i+1] <-  N[i] + r*N[i]*((1-N[i])/K)
}

plot(N, main = "r=0.05")

```

## 4. Measles data in the UK

On Canvas (<https://mycourses.unh.edu/courses/88910/files/6917786?module_item_id=2094185>), I've uploaded historical data on total measles cases and deaths for Wales and the UK. Load the data into R and build figures. Use the data to explain as much as possible about what is happening with the dynamics of measles.

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)

measles <- read.table('measles_wales_and_UK.txt',
                    header=TRUE)

ggplot(measles,aes(x=Year,y=Notifications))+
  geom_line(linewidth=0.7)+
  geom_vline(xintercept=1968, color="red",
             linetype="dashed",linewidth=0.5)+
  geom_vline(xintercept=1988, color="green",
             linetype="dashed",linewidth=0.5)+
  geom_vline(xintercept=1996, color="blue",
             linetype="dashed",linewidth=0.5)+
  theme_minimal()+
  ylab("Measle cases")
```

The above line graph depicts the fluctuations in the numbers of measles cases over the years. Three important years/events are highlighted: red line represents the year 1968, when the first iteration of the measles vaccine was introduced into the UK routine childhood immunization, green represents 1988 with the MMR vaccine introduction, and blue marks the rollout of the two-dose MMR vaccine in 1996 [@oxfordvaccinegroup2024]. The most dramatic decrease occurred right after 1968, from a peak of over 750000 cases to one-fifth of it. The second notable decrease appeared after the MMR vaccine introduction, with the number of cases falling from \~86000 in 1988 to \~26000 in 1989. The count has flatlined since the start of the 90s.

```{r}
ggplot(measles,aes(x=Year,y=Total_Deaths))+
  geom_line(linewidth=0.7)+
  geom_vline(xintercept=1968, color="red",
             linetype="dashed",linewidth=0.5)+
  geom_vline(xintercept=1988, color="green",
             linetype="dashed",linewidth=0.5)+
  geom_vline(xintercept=1996, color="blue",
             linetype="dashed",linewidth=0.5)+
  theme_minimal()+
  ylab("Deaths caused by measles")
```

The second graph depicts the numbers of deaths caused by measles over time. Death had had an exponential decrease since before vaccination, with cases peaking at almost \~1200 in the early 1940s. Since 1968, deaths have dwindled from over 100 to none or only a couple a year. The third graph zooms into changes in the number of deaths from the 1960s to present.

```{r}
measles %>% filter(Year>"1960") %>% 
ggplot(aes(x=Year,y=Total_Deaths))+
  geom_line(linewidth=0.7)+
  geom_vline(xintercept=1968, color="red",
             linetype="dashed",linewidth=0.5)+
  geom_vline(xintercept=1988, color="green",
             linetype="dashed",linewidth=0.5)+
  geom_vline(xintercept=1996, color="blue",
             linetype="dashed",linewidth=0.5)+
  theme_minimal()+
  ylab("Deaths caused by measles")
```

## 5-7. Population genetics

The code below implements a Wright-Fisher model with genetic drift and mutation. Use and modify the code to create plots and answer the following questions.

```{r, echo=TRUE, eval=FALSE}
# Part 1: Wright-Fisher model with genetic drift and mutation

# Created by Easton R. White

# Last edited: 4-Oct-2021

# Setting up parameter values
N=1000      # population size 
p_ini= 0.1   # the initial frequency of the A allele
u = 0.1      # mutation rate from a to A
max.time=100 # time to run simulations


# Custom built function
sample.population <- function(p,N){
  next_pop = vector('character',length = N)
  for (i in 1:N){
    new_allele = sample(x = c('A','a'),size = 1,prob=c(p,1-p))
    next_pop[i] = as.character(new_allele)
  }
  return(sum(next_pop=='A')/N) # returns the value for p
}



p = vector(mode = 'numeric',length = max.time)
p[1]=p_ini

for (t in 1:(max.time-1)){
  p[t+1] = sample.population(p[t],N)
}

plot(1:max.time,p,ylim=c(0,1),type='l',las=1,ylab='Freq(A)',xlab='Time')
```

### Population genetics questions

# 1. What is the effect of population size (N) on the probability of extinction of A allele?

The fixed population size (N=1000) ensures that the A allele extinction probability remain really low. However if we had a smaller population size the probability of extinction of any allele would increase substantially. In summary: the smaller the population, the higher probability of genetic drift and therefore allele fixation. 

<!--# Khanh: it's my understanding that it's really hard for this population to not have any A allele (probability =1/1000?). I don't know anything about genetics so please fix language as well. -->

<!--# Olivia: My understanding of Genetic Drift generally speaking is that the lower the population is the higher chance of allele fixation. Essentially the smaller the population the faster that entire population will have that allele (especially if it's the dominant A instead of the recessive a allele), so I added that second sentence to our answer. Think of like a small group of ~20 people migrates to an island how many generations it would take before the dominant allele took over completely versus if it was a giant group of 20,000 people. -->

# 2. How does the initial frequency of the A allele affect the probability it will reach fixation?

It is possible for the A allele to reach fixation but it will require a lot more simulations as the initial frequency of 0.1 is relatively low (a high p[t] will more likely lead to a slightly higher p[t+1]).

# 3. Let's now study the combined effects of mutation and genetic drift. Modify the above code (specifically the sample.population() function) to include a probability, u, that if a small "a" allele is chosen, it will mutate to be a big A allele. An if/else statement and or the command rbinom() might be helpful. How does this affect our findings in questions 1 and 2?

```{r}
sample.population <- function(p,N){
  next_pop = vector('character',length = N)
  for (i in 1:N){
    new_allele = sample(x = c('A','a'),size = 1,prob=c(p,1-p))
    if (new_allele=='a'){
      new_allele=sample(x=c('A','a'),size=1,prob=c(u,1-u))
    } else {
      new_allele
    }
    next_pop[i] = as.character(new_allele)
  }
  return(sum(next_pop=='A')/N) # returns the value for p
}

p = vector(mode = 'numeric',length = max.time)
p[1]=p_ini

for (t in 1:(max.time-1)){
  p[t+1] = sample.population(p[t],N)
}

plot(1:max.time,p,ylim=c(0,1),type='l',las=1,
     ylab='Freq(A)',xlab='Time')
```

The mutation drastically decreases the probability of A alleles becoming extinct and accelerated the process of A allele fixation.
